\begin{abstract}
Given a graph $G = (V, E)$, 
a coloring function $\chi: V \rightarrow C$, 
assigning each vertex a color,
is called \emph{convex} if, 
for every color $c \in C$, 
the set of vertices with color $c$ induces a connected subgraph of $G$.
%
In the \textsc{Convex Recoloring} problem a colored graph $G_\chi$ is given, 
and the goal is to find a convex coloring $\chi'$ of 
$G$ that \emph{recolors} a minimum number of vertices.
%
The \textsc{2-Convex Recoloring} problem (\TWOCR{}) is the special case,
in which the given coloring $\chi$ assigns the same color to at most
two vertices.  
\TWOCR{} is known to be NP-hard even if $G$ is a path.

We show that the weighted \TWOCR{} problem can not be approximate within any 
ratio on general graphs.  

We provide an alternative definition of \TWOCR{} in terms of maximum independent set of paths,
which leads to a natural greedy algorithm.  
We prove that its approximation ratio is $\frac{3}{2}$, 
and show that this analysis is tight.
%
This is the first constant factor approximation algorithm for a variant of 
\CRP{} in general graphs.
%
For the special case, 
where $G$ is a path, 
the algorithm obtains a ratio of $\frac{5}{4}$, 
an improvement over the previous best known approximation.
%
We also consider the problem of determining whether a given graph 
has a convex recoloring of size $k$.  
We use the above mentioned characterization of 
\TWOCR{}{} to show that a problem kernel of size $4k$ can be obtained
in linear time,
and to design a $2^{O(k \log k)}$ time algorithm for parametrized \TWOCR{}.
\end{abstract}
